Generalized spaces for constructive algebra

### 2019-09-23 slides-herrsching2019.tex
Some notes slides

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}

@@

\begin{frame}{Frames presented by theories}
  \justifying
  \textbf{Definition.} A \hil{geometric theory} consists of
  \begin{enumerate}
    \small
    \item a set of sorts: $X$, $Y$, $Z$, \ldots
    \item a set of function symbols: $f : X \times Y \to Z$, \ldots
    \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
    \item a set of axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
  \end{enumerate}
  % A geometric signature is \hil{propositional} iff its set
  % of sorts is empty.

@@

  \jnote{1}{
    A geometric theory is called \emph{propositional} iff its set of
    sorts is empty. Hence the only ingredients of a propositional geometric
    theory are a set of nullary relation symbols (atomic propositions) and a
    set of axioms.

    Associated to any propositional geometric theory~$\TT$ is its
    \emph{Lindenbaum algebra}. This is the set of formulas over the signature
    of the theory modulo provable equivalence. Endowed with the partial order
    given by
    \[ [\varphi] \leq [\psi] \quad\vcentcolon\Longleftrightarrow\quad
      \text{$\TT$ proves $\varphi \vdash \psi$}, \]
    the Lindenbaum algebra becomes a frame.

    The Lindenbaum algebra of~$\TT$ is the free frame generated by the
    atomic propositions of~$\TT$ modulo the axioms of~$\TT$. The associated
    locale~$L(\TT)$ is the \emph{classifying locale of~$\TT$}. Its points are
    in canonical bijection with the set-based models of~$\TT$.
  }
\end{frame}

@@

  \textbf{Non-example.} The \hil{presheaf of constant
  functions},~$\C_{\text{c}}(U) = \text{``the set of constant functions~$U \to
  \RR$''}$, it usually not a sheaf.
\end{frame}

@@

  \jnote{1}{
    A \emph{formula over an open~$U$} of~$X$ is a first-order formula (made up
    using~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\ {\Rightarrow}\ {\forall}\
    {\exists}$) over the signature which has one sort for each sheaf~$F$, one
    constant symbol of sort~$F$ for each section~$s \in F(U)$, one function
    symbol~$f : F \to G$ for each morphism of sheaves, and so on.
  }
\end{frame}

>>>>

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
%\newcommand{\jnote}[2]{}

@@

\begin{frame}{Frames presented by theories}
  \justifying
  \textbf{Definition.} A \hil{geometric theory} consists of
  \begin{enumerate}
    \small
    \item a set of sorts: $X$, $Y$, $Z$, \ldots
    \item a set of function symbols: $f : X \times Y \to Z$, \ldots
    \item a set of relation symbols: $R \hookrightarrow X \times Y \times Z$, \ldots
    \item a set of axioms: $\varphi \vdash_{x:X, y:Y} \psi$, \ldots
  \end{enumerate}

@@

  \jnote{1}{
    The formulas appearing to the left and the right of the turnstile in axioms
    of a geometric theory have to be \emph{geometric formula}. A formula is
    \emph{geometric} iff it is built using only~${=}\ {\top}\ {\wedge}\
    {\bot}\ {\vee}\ {\bigvee}\ {\exists}$ (but no ${\Rightarrow}\
    {\forall}$). Universal quantification and implication can to some extent be
    emulated by using free variables or the turnstile; but this allows
    universal quantification and implication only to be used once, on top
    level, not in nested subformulas.

    A geometric theory is called \emph{propositional} iff its set of
    sorts is empty. Hence the only ingredients of a propositional geometric
    theory are a set of nullary relation symbols (atomic propositions) and a
    set of axioms.
  }

  \jnote{2}{
    \textbf{Example.} The theory of rings has:
    \begin{enumerate}
      \item one sort: $R$
      \item five constant symbols: $0$ and $1$ (nullary), $-$ (unary), $+$ and
      $\cdot$ (binary)
      \item no relation symbols
      \item the usual axioms, such as $\top \vdash_{x:R,y:R} x + y = y + x$
    \end{enumerate}

    \textbf{Example.} The theory of surjections~$\NN \twoheadrightarrow \RR$ has:
    \begin{enumerate}\justifying
      \item no sorts
      \item no function symbols
      \item several nullary relation symbols, one for each pair~$\langle n,x
      \rangle \in \NN \times \RR$, written~``$\varphi_{nx}$''
      \item axioms: the axiom~$\top \vdash \bigvee_{x \in \RR} \varphi_{nx}$
      for each~$n \in \NN$; the axiom~$\varphi_{nx} \wedge \varphi_{ny} \vdash
      \bigvee\{ \top \,|\, x = y \}$ for each~$n \in \NN$, $x,y \in \RR$ (the
      disjunction appearing in that axiom is over a subsingleton of formulas);
      the axiom~$\top \vdash \bigvee_{n \in \NN} \varphi_{nx}$ for each~$x \in
      \RR$
    \end{enumerate}
  }

  \jnote{3}{
    Associated to any propositional geometric theory~$\TT$ is its
    \emph{Lindenbaum algebra}. This is the set of formulas over the signature
    of the theory modulo provable equivalence. Endowed with the partial order
    given by
    \[ [\varphi] \leq [\psi] \quad\vcentcolon\Longleftrightarrow\quad
      \text{$\TT$ proves $\varphi \vdash \psi$}, \]
    the Lindenbaum algebra becomes a frame.

    The Lindenbaum algebra of~$\TT$ is the free frame generated by the
    atomic propositions of~$\TT$ modulo the axioms of~$\TT$. The associated
    locale~$L(\TT)$ is the \emph{classifying locale of~$\TT$}. Its points are
    in canonical bijection with the set-based models of~$\TT$.

    \textbf{Definition.} The \emph{localic real line} is the classifying locale
    of the theory of Dedekind cuts.
    The \emph{locale of surjections~$\NN \twoheadrightarrow \RR$} is the
    classifying locale of the theory of surjections~$\NN \twoheadrightarrow
    \RR$.
    The \emph{empty locale} is the classifying locale of the \emph{inconsistent
    theory} (which has no sorts, no function symbols, no relations symbols, and
    one axiom, namely~$\top \vdash \bot$).

    The points of the localic real line are in canonical bijection with
    the Dedekind cuts, hence with the elements of the set of (Dedekind) real
    numbers. The points of the locale of surjections~$\NN \twoheadrightarrow
    \RR$ are in canonical bijection with the surjections~$\NN
    \twoheadrightarrow \RR$. Assuming a classical metatheory, there are no such
    points, but the locale is still nontrivial.
  }

  \jnote{4}{
    Any locale is (isomorphic to) the classifying locale of some propositional
    geometric theory, namely the ``theory of its points''. The same locale can
    often be presented by widely different-looking theories, and the
    topos-theoretic version of this observation is the basis for Olivia
    Caramello's research program.

    \textbf{Exercise.} Which explicit theory is the two-point locale (the
    locale induced from the discrete topological space~$\{0,1\}$) the
    classifying locale of?
  }
\end{frame}

@@

  \textbf{Non-example.} The \hil{presheaf of constant
  functions},~$\C_{\text{c}}(U) = \text{``the set of constant functions~$U \to
  \RR$''}$, is usually not a sheaf.

  \jnote{1}{
    All definitions on this slide literally also make sense for topological
    spaces instead of locales, since only the notion of opens and their
    inclusion relation is used.

    The maps~$(\cdot)|^U_V$ are called \emph{restriction maps}. For the
    sheaf~$\C$ of continuous real-valued functions, they are given by actual
    restriction of functions to smaller domains, that is by
    \[ \C(U) \longrightarrow \C(V),\ s \longmapsto s|_V. \]

    \textbf{Exercise.} Find, in the case~$X = [0,1]$, a compatible family of
    sections of the presheaf~$\C_\text{c}$ which shows that this
    presheaf is not a sheaf.
  }
\end{frame}

@@

  \jnote{1}{
    A \emph{formula over an open~$U$} of~$X$ is a first-order formula (made up
    using~${=}\ {\top}\ {\wedge}\ {\bot}\ {\vee}\ {\Rightarrow}\ {\forall}\
    {\exists}$) over the signature which has one sort for each sheaf~$F$, one
    constant symbol of sort~$F$ for each section~$s \in F(U)$, one function
    symbol~$f : F \to G$ for each morphism of sheaves, and so on.

    \textbf{Proposition.} The sheaf semantics is monotone and local in the
    following sense: If~$V \leq U$, then~$U \models \varphi$ implies~$V \models
    \varphi$. If~$U = \bigvee_i U_i$ and if~$U_i \models \varphi$ for all~$i$,
    then~$U \models \varphi$.

    \textbf{Exercise.} For any formula~$\varphi$ over any open~$U$ of any
    locale~$X$, there is a largest open~$V \leq U$ such that~$V \models
    \varphi$. (Hint: Set~$V \defeq \bigvee\{ W \leq U \,|\, W \models \varphi \}$
    and exploit the monotonicity and locality of the sheaf semantics.)

    \textbf{Exercise.} For any formula~$\varphi$ over any open~$U$ of any
    locale~$X$, $U \models \neg\neg\varphi$ iff there exists an open~$V \leq U$
    which is dense in~$U$ such that~$V \models \varphi$.
  }
\end{frame}
